---
title: "ADS1 analysis - Cronicas study"
author: "Josiah Kephart"
date: "October 18, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Project outline

Peru, along with other low and middle-income countries (LMICs) is experience a transition in the burden of disease from infectious diseases to chronic diseases. Air pollution, both ambient and indoor, is a major risk factor for chronic disease of the lungs and heart. Air pollution is known to vary by geography due to differences in both regional development and individual behaviors. Characterizing the airborne exposures faced by Peruvians in both urban and rural areas is critical to understanding the contribution of air pollution to rising chronic disease and informing environmental regulations and household interventions to reduce exposures.

To accomplish this aim, a study of household air pollution and clinical outcomes, known as, CRONICAS, was conducted in Lima and Puno, Peru in 2014-2015. 

48 hour direct-reading measurements of PM2.5 and CO at one minute temporal resolution were taken in roughly 400 households in Lima (urban) and 400 households in Puno (rural) who used traditional, biomass stoves for cooking.

###Research question
How do household PM2.5 levels compare between urban and rural areas of Peru?

```{r urban, include = FALSE}

library(data.table)
library(dplyr)
library(plyr)
library(stringi)
library(ggplot2)

setwd("C:/Users/josia/Google Drive/Research/CRONICAS")
setwd("./Data/Urban/MedicionesUrbano/MedicionesUrbano/PDR URBANO/Data")

urban_samples <- list.files(path = ".", pattern = ".CSV")

##Add household ID to data
read_csv_filename <- function(urban_samples){
  ret <- fread(urban_samples, showProgress = T, autostart = 15)
  ret$Source <- urban_samples 
  ret
}

samples_urban <- ldply(urban_samples, read_csv_filename) 

##Clean up columns

urbandata <- samples_urban
urbandata$pmurban <- rowSums(urbandata[,c("Avg.(mg/m³)", "Avg.(mg/m3)")], na.rm = T)
urbandata$hid <- stri_sub(urbandata$Source,7,12)
urbandata$yr <- stri_sub(urbandata$Source, 19,20)
urbandata$ndate <- paste(urbandata$Date,urbandata$yr)
urbandata$ndatetime <- paste(urbandata$ndate, urbandata$Time)
urbandata$datetime <- lubridate::dmy_hms(urbandata$ndatetime)
urbandata$hour <- stri_sub(urbandata$Time,1,2)
urbandata$hidtime <- paste(urbandata$hid,urbandata$ndate,urbandata$hour, sep = "_")
urbandata$hiddt <- gsub(" ", "", urbandata$hidtime)
urbandata$strtdate <- stri_sub(urbandata$Source,14,20)
urbandata <- select_(urbandata, "hiddt", "pmurban","strtdate", "hid","datetime")
head(urbandata)

quant <- "quantile(pmurban, probs=0.95)"
funs <- c("mean", "max", "median",quant)

sumdata_urban <- urbandata %>%
  group_by(hiddt) %>%
  summarise_at(vars(pmurban), funs) %>% 
  mutate(hid = stri_sub(hiddt, 1,6)) %>%
  mutate(hr = stri_sub(hiddt, -2,-1)) 
  
head(sumdata_urban)
  

#create columns in summarized data set 


unihid_urban <- unique(sumdata_urban$hid)
length(urban_samples)
length(unihid_urban)

urbanpm <- sumdata_urban %>%
  summarise(mean = mean(mean)) 
  
urbanpm$median <- sumdata_urban %>%
  summarise(median = mean(median)) 

urbanpm$max95 <- sumdata_urban %>%
  summarise(max95 = max(quantile))

urbanpm$sd <- sumdata_urban %>%
  summarise(sd = sd(mean))

```

```{r rural, include=FALSE}

setwd("C:/Users/josia/Google Drive/Research/CRONICAS/Data/Rural/MedicionesRural/MedicionesRural/PDR RURAL/Data")

rural_samples <- list.files(path = ".", pattern = ".CSV")

##Add household ID to data
read_csv_filename <- function(rural_samples){
  ret <- fread(rural_samples, showProgress = T, autostart = 15)
  ret$Source <- rural_samples #EDIT
  ret
}

samples_rural <- ldply(rural_samples, read_csv_filename); 

##Clean up columns

ruraldata <- samples_rural
ruraldata$pmrural <- rowSums(ruraldata[,c("Avg.(mg/m³)", "Avg.(mg/m3)")], na.rm = T)
ruraldata$hid <- stri_sub(ruraldata$Source,7,12)
ruraldata$yr <- stri_sub(ruraldata$Source, 19,20)
ruraldata$ndate <- paste(ruraldata$Date,ruraldata$yr)
ruraldata$ndatetime <- paste(ruraldata$ndate, ruraldata$Time)
ruraldata$datetime <- lubridate::dmy_hms(ruraldata$ndatetime)
ruraldata$hour <- stri_sub(ruraldata$Time,1,2)
ruraldata$hidtime <- paste(ruraldata$hid,ruraldata$ndate,ruraldata$hour, sep = "_")
ruraldata$hiddt <- gsub(" ", "", ruraldata$hidtime)
ruraldata$strtdate <- stri_sub(ruraldata$Source,14,20)
ruraldata <- select_(ruraldata, "hiddt", "pmrural","strtdate", "hid","datetime")
head(ruraldata)

#Create summarized data set by ID & hour

quant <- "quantile(pmrural, probs=0.95)"
funs <- c("mean", "max", "median",quant)

sumdata_rural <- ruraldata %>%
  group_by(hiddt) %>%
  summarise_at(vars(pmrural), funs) %>% 
  mutate(hid = stri_sub(hiddt, 1,6)) %>%
  mutate(hr = stri_sub(hiddt, -2,-1)) 
  
head(sumdata_rural)

#create columns in summarized data set 

unihid_rural <- unique(sumdata_rural$hid)
length(samples_rural)
length(unihid_rural)

ruralpm <- sumdata_rural %>%
  summarise(mean = mean(mean)) 
  
ruralpm$median <- sumdata_rural %>%
  summarise(median = mean(median)) 

ruralpm$max95 <- sumdata_rural %>%
  summarise(max95 = max(quantile))

ruralpm$sd <- sumdata_rural %>%
  summarise(sd = sd(mean))

```
#Exploratory Data Analysis
###Number of Urban vs Rural samples

Urban samples: 336

Rural samples: 357


Unique households in rural samples: 319

###Summary statistics of hourly PM concentrations
####Urban PM2.5
```{r summarystats, echo=FALSE}
urbanpm
```
####Rural PM2.5
```{r summarystats2, echo=FALSE}
ruralpm
```


```{r plotsmean, echo=FALSE}
hrurban <- sumdata_urban %>%
  group_by(hr) %>%
  summarise_at(vars(median),mean)

hrrural <- sumdata_rural %>%
  group_by(hr) %>%
  summarise_at(vars(median),mean)

#Plot mean hourly household median

urbanmeanbase <- ggplot() + 
  geom_line(data = hrurban, aes(hr,median, group = 1, color = "Urban households"), colour = 4) +
  ylab("Median PM2.5 mg/m3") + 
  xlab("Daily Hour") + 
  geom_hline(yintercept = 0.035) +   
  ggtitle("Mean hourly median PM2.5 concentrations in urban households") 
```

```{r ploturbanmean, echo=FALSE}
urbanmeanbase
```

```{r save1, echo=FALSE}
ggsave("./urbanmean.jpg", plot=last_plot())
```

```{r plotsmedian2, echo=FALSE}
urbanmeanbase + geom_line(data = hrrural, aes(hr,median, group = 1, color = "Rural households")) +
  ggtitle("Mean hourly median PM2.5 concentrations in urban and rural households") 
ggsave("./bothmean.jpg", plot=last_plot())
```

```{r plotsmedian3, echo=FALSE}

#Create log10 scaled graph

log10ug_urb <- hrurban %>%
  mutate(log10urban = log10(median*1000)) %>%
  select(log10urban, hr)

log10ug_rur <- hrrural %>%
  mutate(log10rural = log10(median*1000)) %>%
  select(log10rural, hr)

logurb_base <- ggplot() + 
  geom_line(data = log10ug_urb, aes(hr,log10urban, group = 1, color = "Urban households"), colour = 4) +
  ylab("Log10 PM2.5 ug/m3") + 
  xlab("Daily Hour") + 
  geom_hline(yintercept = log10(0.035*1000), col="black") +   
  ggtitle("Mean hourly median PM2.5 concentrations in urban households") 
logurb_base

ggsave("./logurbanmean.jpg", plot=last_plot())
```

```{r plotsmedian4, echo=FALSE}
logurb_base + geom_line(data = log10ug_rur, aes(hr,log10rural, group = 1, color = "Rural households")) +
  ggtitle("Mean hourly median PM2.5 concentrations in urban and rural households") 
ggsave("./logbothmean.jpg", plot=last_plot())

```

```{r boxplots, echo=FALSE}

#urban box plots

sumdata_urban_n <- sumdata_urban %>%
  mutate(hr_num = as.numeric(hr))

#jpeg('urbanbox.jpg')
boxplot(median ~ hr, data=sumdata_urban_n, main="Median hourly PM2.5 in urban households", xlab="Hour of day",ylab="PM2.5 mg/m3", col="blue")
abline(h=0.35, col="green")
dev.copy(jpeg,'urbanbox.jpg')
dev.off()

#urban log10

sumdata_urban_log <- sumdata_urban %>%
  mutate(logurban = log10(median*1000+1))

boxplot(logurban ~ hr, data=sumdata_urban_log, main="Median hourly PM2.5 in urban households", xlab="Hour of day",ylab="Log10 PM2.5 ug/m3")
abline(h=log10(0.35*1000), col="green")
dev.copy(jpeg,'logurbanbox.jpg')
dev.off()

#rural box plots

sumdata_rural_n <- sumdata_rural %>%
  mutate(hr_num = as.numeric(hr))

boxplot(median ~ hr, data=sumdata_rural_n, main="Median hourly PM2.5 in rural households", xlab="Hour of day",ylab="PM2.5 mg/m3")
abline(h=0.35, col="green")
dev.copy(jpeg,'ruralbox.jpg')
dev.off()

#rural log10

sumdata_rural_log <- sumdata_rural %>%
  mutate(logrural = log10(median*1000+1))

boxplot(logrural ~ hr, data=sumdata_rural_log, main="Median hourly PM2.5 in rural households", xlab="Hour of day",ylab="Log10 PM2.5 ug/m3")
abline(h=log10(0.35*1000), col="green")
dev.copy(jpeg,'logruralbox.jpg')
dev.off()


```