---
title: "ADS1 outline - Cronicas study"
author: "Josiah Kephart"
date: "October 4, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Project outline

Cronicas is a study of household air pollution and clinical outcomes in Lima and Puno, Peru in 2014-2015. 

48 hour direct-reading measurements of PM2.5 and CO at one minute temporal resolution were taken in roughly 400 households in Lima (urban) and 400 households in Puno (rural) who used traditional, biomass stoves for cooking.

###Research question
How do household PM2.5 levels compare between urban and rural areas of Peru?

```{r urban, include = FALSE}

library(data.table)
library(dplyr)
library(plyr)
library(stringi)
library(ggplot2)

setwd("C:/Users/josia/Google Drive/Research/CRONICAS")
setwd("./Data/Urban/MedicionesUrbano/MedicionesUrbano/PDR URBANO/Data")

urban_samples <- list.files(path = ".", pattern = ".CSV")

##Add household ID to data
read_csv_filename <- function(urban_samples){
  ret <- fread(urban_samples, showProgress = T, autostart = 15)
  ret$Source <- urban_samples #EDIT
  ret
}

samples_urban <- ldply(urban_samples, read_csv_filename); 

##Clean up columns

testdata <- samples_urban
testdata$pmurban <- rowSums(testdata[,c("Avg.(mg/m³)", "Avg.(mg/m3)")], na.rm = T)
testdata$hid <- stri_sub(testdata$Source,7,12)
testdata$yr <- stri_sub(testdata$Source, 19,20)
testdata$ndate <- paste(testdata$Date,testdata$yr)
testdata$ndatetime <- paste(testdata$ndate, testdata$Time)
testdata$datetime <- lubridate::dmy_hms(testdata$ndatetime)
testdata$hour <- stri_sub(testdata$Time,1,2)
testdata$hidtime <- paste(testdata$hid,testdata$ndate,testdata$hour, sep = "_")
testdata$hiddt <- gsub(" ", "", testdata$hidtime)
testdata$strtdate <- stri_sub(testdata$Source,14,20)
testdata <- select_(testdata, "hiddt", "pmurban","strtdate", "hid","datetime")
head(testdata)

quant <- "quantile(pmurban, probs=0.95)"
funs <- c("mean", "max", "median",quant)

sumdata_urban <- testdata %>%
  group_by(hiddt) %>%
  summarise_at(vars(pmurban), funs) %>% 
  mutate(hid = stri_sub(hiddt, 1,6)) %>%
  mutate(hr = stri_sub(hiddt, -2,-1)) 
  
head(sumdata_urban)
  
#create columns in summarized data set 

head(sumdata_urban)  

unihid_urban <- unique(sumdata_urban$hid)
length(urban_samples)
length(unihid_urban)

urbanpm <- sumdata_urban %>%
  summarise(mean = mean(mean)) 
  
urbanpm$median <- sumdata_urban %>%
  summarise(median = mean(median)) 

urbanpm$max95 <- sumdata_urban %>%
  summarise(max = max(quantile))

hrly <- sumdata_urban %>%
  group_by(hr) 
```

```{r rural, include=FALSE}

setwd("C:/Users/josia/Google Drive/Research/CRONICAS/Data/Rural/MedicionesRural/MedicionesRural/PDR RURAL/Data")

rural_samples <- list.files(path = ".", pattern = ".CSV")

##Add household ID to data
read_csv_filename <- function(rural_samples){
  ret <- fread(rural_samples, showProgress = T, autostart = 15)
  ret$Source <- rural_samples #EDIT
  ret
}

samples_rural <- ldply(rural_samples, read_csv_filename); 

##Clean up columns

testdata <- samples_rural
testdata$pmrural <- rowSums(testdata[,c("Avg.(mg/m³)", "Avg.(mg/m3)")], na.rm = T)
testdata$hid <- stri_sub(testdata$Source,7,12)
testdata$yr <- stri_sub(testdata$Source, 19,20)
testdata$ndate <- paste(testdata$Date,testdata$yr)
testdata$ndatetime <- paste(testdata$ndate, testdata$Time)
testdata$datetime <- lubridate::dmy_hms(testdata$ndatetime)
testdata$hour <- stri_sub(testdata$Time,1,2)
testdata$hidtime <- paste(testdata$hid,testdata$ndate,testdata$hour, sep = "_")
testdata$hiddt <- gsub(" ", "", testdata$hidtime)
testdata$strtdate <- stri_sub(testdata$Source,14,20)
testdata <- select_(testdata, "hiddt", "pmrural","strtdate", "hid","datetime")
head(testdata)

quant <- "quantile(pmrural, probs=0.95)"
funs <- c("mean", "max", "median",quant)

sumdata_rural <- testdata %>%
  group_by(hiddt) %>%
  summarise_at(vars(pmrural), funs) %>% 
  mutate(hid = stri_sub(hiddt, 1,6)) %>%
  mutate(hr = stri_sub(hiddt, -2,-1)) 
  
head(sumdata_rural)
  
#create columns in summarized data set 

head(sumdata_rural)  

unihid_rural <- unique(sumdata_rural$hid)
length(samples_rural)
length(unihid_rural)

ruralpm <- sumdata_rural %>%
  summarise(mean = mean(mean)) 
  
ruralpm$median <- sumdata_rural %>%
  summarise(median = mean(median)) 

ruralpm$max95 <- sumdata_rural %>%
  summarise(max = max(quantile))

hrly <- sumdata_rural %>%
  group_by(hr) 

```
#Exploratory Data Analysis
###Number of Urban vs Rural samples

N of urban samples: 336
N of rural samples: 357

###Number of unique households Urban vs Rural

Unique households in urban samples: 287
Unique households in rural samples: 319

###Average, median, 95th percentile of hourly PM concentrations
```{r summarystats}
urbanpm
ruralpm
```


```{r plots, echo=FALSE}
hrurban <- sumdata_urban %>%
  group_by(hr) %>%
  summarise_at(vars(median),mean)

hrrural <- sumdata_rural %>%
  group_by(hr) %>%
  summarise_at(vars(median),mean)

base <- ggplot() + 
  geom_line(data = hrurban, aes(hr,median, group = 1, color = "Urban households")) +
  ylab("Mean PM2.5 mg/m3") + xlab("Daily Hour") + labs(color="PDR pilots") +
  geom_hline(yintercept = 0.035) +   
  ggtitle("Mean hourly median PM2.5 concentrations in urban households") 

base
```

```{r plots2, echo=FALSE}
base + geom_line(data = hrrural, aes(hr,median, group = 1, color = "Rural households")) +
  ggtitle("Mean hourly median PM2.5 concentrations in urban and rural households") 
```